IRT procedures for shortening tests

### 2024-10-09 the-good-the-bad-the-ugly.qmd
a lot

---
title: "IRT procedures for shortening tests"
author: "Ottavia M. Epifania & Friends"
format: 
  revealjs: 
    theme: mytheme.scss
background-opacity: "0.45"
title-slide-attributes: 
   data-background-image: "https://psicostat.github.io/shine-bright/gbu.jpg"
   data-background-size: 350px, cover
   data-background-position: 50% 5%, center
---

```{r, setup, include=FALSE}
library(knitr)

@@

## Item Locating Algotihm -- ILA
### The bad

@@

```{r}
#| fig-align: center
knitr::include_graphics("img/ila.png")
```

@@

[^1]: **FRANK**ly I didnt' know how to name it

## Bruto 
### The bad

## Bullets

When you click the **Render** button a document will be generated that includes:

-   Content authored with markdown
-   Output from executable code

## Code

When you click the **Render** button a presentation will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
1 + 1
```

>>>>

---
title: "IRT procedures for shortening tests"
author: "Ottavia M. Epifania & Friends"
format: 
  revealjs: 
    theme: mytheme.scss
background-opacity: "0.45"
title-slide-attributes: 
   data-background-image: "https://psicostat.github.io/shine-bright/gbu.jpg"
   data-background-size: 350px, cover
   data-background-position: 50% 5%, center
server: shiny
---

```{r, setup, include=FALSE}
library(knitr)
library(shiny)
IRT <- function(theta, a = 1, b = 0, c = 0,e = 1) {
  y <- c + (e - c) * exp(a * (theta - b)) / (1 + exp(a * (theta - b)))
  y[is.na(y)] = 1
  return(y)
}

i_info <- function(b, a=1,c=0, theta = seq(-5,5,length.out=1000)){

P <- NULL 
Q <- NULL
Ii <- NULL
for(i in 1:1000){
  P[i] <- 1/(1+ exp (-a*(theta[i] - b)))
  Q[i]= 1-P[i]
  Ii[i] =(a*Q[i]*(P[i]-c)^2)/(P[i]*((1-c)^2)) # (3PL)
   }
return(Ii)
}
# Function to get all item information
item_info <- function(b,a=1, c= 0){
item <- NULL
  for(i in 1:length(b)){
  item[[i]] <- i_info(b[i],a[i])
  }
return(item)
}

set.seed(999)
```

## The 2-Parameter Logistic Model

$$P(x_{pi} = 1|\theta_p, b_i, a_i) = \frac{\exp[a_i(\theta_p - b_i)])}{1 + \exp[a_i(\theta_p - b_i)])}$$

$$IIF_i = a_i^2P(\theta, b_i, a_i)[1-P(\theta, b_i, a_i)]$$

$$TIF = \sum_{i = 1}^{||B||} IIF_i$$

## Intuitively

```{r}
#| panel: sidebar
tags$style(HTML(".js-irs-0 .irs-single, .js-irs-0 .irs-bar-edge, .js-irs-0 .irs-bar {background: royalblue}"))
tags$style(HTML(".js-irs-1 .irs-single, .js-irs-1 .irs-bar-edge, .js-irs-1 .irs-bar {background: magenta}"))
tags$style(HTML(".js-irs-2 .irs-single, .js-irs-2 .irs-bar-edge, .js-irs-2 .irs-bar {background: seagreen}"))
tags$style(HTML(".js-irs-3 .irs-single, .js-irs-3 .irs-bar-edge, .js-irs-3 .irs-bar {background: royalblue}"))
tags$style(HTML(".js-irs-4 .irs-single, .js-irs-3 .irs-bar-edge, .js-irs-4 .irs-bar {background: magenta}"))
tags$style(HTML(".js-irs-5 .irs-single, .js-irs-5 .irs-bar-edge, .js-irs-5 .irs-bar {background: seagreen}"))

 sliderInput("b1", "b1",
                  min = -3, max = 3,
                  value = 0, step = 0.1)
sliderInput("b2", "b2",
                  min = -3, max = 3,
                  value = 0, step = 0.1)
sliderInput("b3", "b3",
                  min = -3, max = 3,
                  value = 0, step = 0.1)

sliderInput("a1", "a1",
                  min = 0.20, max = 3,
                  value = 1, step = 0.1)
sliderInput("a2", "a2",
                  min = 0, max = 3,
                  value = 0.70, step = 0.1)
sliderInput("a3", "a3",
                  min = 0, max = 3,
                  value = 1.90, step = 0.1)
checkboxInput("showTif", "Show Test Information Function", value = FALSE)

@@

```{r}
#| panel: fill

plotOutput("plot", height = "300px")

plotOutput("tif", height = "300px")
```

```{r}
#| context: server
 output$plot <- renderPlot({
   theta <- theta <- seq(-7, 7, .001)
  par(mar = c(5,7,4,2) + 0.1)
  b = c(input$b1, input$b2, input$b3)
  a = c(input$a1, input$a2, input$a3)

 plot(theta, IRT(theta, b = input$b1, a = input$a1), 
                 type = "l", lwd = 2, lty = 2,
     col = "royalblue", main = "Item Characteristics Curves - ICCs", 
     ylab = expression(paste("P(", x[p][i],  "= 1|", theta[p], ", ", b[i], ", ", a[i], ")")), 
     xlab = expression(theta), 
     cex.lab = 2.4, 
     cex.main = 3, 
     cex.text=2, cex.axis=1.5)

segments(min(theta)-3, 0.5, 
         input$b1, 0.5, 
         col = "gray56", lty = 3, lwd = 1.3)
segments(input$b1, -0.5, 
         input$b1, 0.5, 
         col = "royalblue", lty = 3, lwd = 1.3)

lines(theta, IRT(theta, b=input$b2,
                 a=input$a2),
       type = "l", lwd = 2, lty = 2,
     col = "magenta")
 lines(theta, IRT(theta, b=input$b3,
                 a=input$a3),
       type = "l", lwd = 2, lty = 2,
     col = "seagreen")
 segments(min(theta)-3, 0.5, 
         input$b2, 0.5, 
         col = "gray56", lty = 3, lwd = 1.3)
segments(input$b2, -0.5, 
         input$b2, 0.5, 
         col = "magenta", lty = 3, lwd = 1.3)
 segments(min(theta)-3, 0.5, 
         input$b3, 0.5, 
         col = "gray56", lty = 3, lwd = 1.3)
segments(input$b3, -0.5, 
         input$b3, 0.5, 
         col = "seagreen", lty = 3, lwd = 1.3)
  })

output$tif <- renderPlot({

   theta <- theta <- seq(-7, 7, .001)
  par(mar = c(5,7,4,2) + 0.1)
  b = c(input$b1, input$b2, input$b3)
  a = c(input$a1, input$a2, input$a3)
c <- item_info(b,a)
Theta <- matrix(seq(-4,4, length.out=1000))
check <- data.frame(Theta,
                    item_info = c[[1]],
                    item_info2 = c[[2]],
                    item_info3 = c[[3]])
d <- do.call('cbind',c)
sum_info1 <- rowSums(d)
d1 <- do.call('cbind',c)
sum_info2 <- rowSums(d1)
plot(check$Theta, check$item_info, cex.lab= 2, main = "Item Information Functions - IIFs",
      cex.lab = 2.4, 
     cex.main = 3, 
     cex.text=2, cex.axis=1.5,
        xlab = expression(theta), ylab = expression(paste("IIF"[i])),
 type = "l", lwd =2,
      col = "royalblue", ylim = c(0,1))
 lines(check$Theta, check$item_info2, lwd =2,
       col = "magenta", lty = 4)
 lines(check$Theta, check$item_info3, lwd =2,
       col = "seagreen", lty = 2)

checkTif <- data.frame(Theta, sum_info1, sum_info2)

if (input$showTif == TRUE) {
  plot(checkTif$Theta, checkTif$sum_info2,
       type = "l", lwd =2,
     col = "black", ylim = c(0, 2),
       xlab = expression(theta), 
     ylab = expression(paste("I(", theta, ")")), 
     cex.axis=1.5, main = "IIFs and Test Information Functions",  
     cex.lab = 2.4, 
     cex.main = 3, 
     cex.text=2)
  lines(check$Theta, check$item_info, lwd =2,
       col = "royalblue", lty = 4)
   lines(check$Theta, check$item_info2, lwd =2,
       col = "magenta", lty = 4)
 lines(check$Theta, check$item_info3, lwd =2,
       col = "seagreen", lty = 2)
}

})
```

## Some premises

::: {.callout-note}
## Key concept

<font size = "4">
**TIF target** ($\mathbf{TIF}^*$) describing the desired characteristics of a test
</font>
:::

::: {.callout-tip}
## The aim of the algorithms

<font size = "4">
*Minimize the distance* between $\mathbf{TIF}^*$ and that of the short test forms (STFs) 
</font>
:::

::: {.callout-warning}
## Their differences

<font size = "4">
The method for selecting and including the items in $Q \subset B$ from the item bank
</font>
:::

::: {.callout-important}
## Mean TIF

<font size = "4">
TIF is considered as mean TIF $\rightarrow$ as the number of items increases, the TIF increases
</font>
:::

## Bruto (tu quoque?!)
### The bad

<font size = "5">

$Q_m \subset B = 2^{||B||}-2$ STFs of different lengths considering different item combination can be developed from $B$. That's what Bruto does

1. $\mathbf{TIF}^{Q_m} =  \frac{\sum_{i \in Q_m} IIF_i}{||Q_m||}$
2. $\overline{\Delta}_{\mathbf{TIF}^{Q_m}} =  \mathit{mean}(|\mathbf{TIF}^* - \mathbf{TIF}^{Q_m}|)$

$Q_{bruto} = \arg \min_{Q_m \subset B} \overline{\Delta}_{\mathbf{TIF}^{Q_m}}$
</font>

```{r}
#| fig-align: center
#| out-width: "50%"
knitr::include_graphics("www/eeato.jpg")
```

## Item Locating Algorithm -- ILA
### The ugly

@@

```{r}
#| fig-align: center
#| out-width: "200%"
knitr::include_graphics("www/ila.png")
```

@@

At $k = 0$, $Q^0 = \emptyset$. For $k \geq 1$, iterate

1. $Q^{k+1} = Q^k \cup \arg \min |TIF^* - IIF_{i \in B \setminus Q^k}|$

2. $TIF_{i \in B\setminus Q^{k+1}}^{k+1} = \frac{IIF_{i \in Q^{k+1}} + IIF_{i \in B \setminus Q^{k+1}}}{||Q^{k+1}|| + 1}$

Until the termination criterion $|TIF^* - TIF^{k+1}| \geq |TIF^* - TIF^{k}|$ is true $\rightarrow$ $Q_{frank} = Q^{k+1}$

[^1]: <font size="3"><font size="6">**FRANK**</font>ly I didnt' know how to name it</font>

## Simulation time

1. Generate an item bank $B$ of $6$ items: 
			- Difficulty parameters: $\mathcal{U}(-3, 3)$
			- Discrimination parameters:  $\mathcal{U}(.90, 2.0)$

2. Random item selections of lengths $l$ from $B$ ($M_l = 3.34 \pm 1.13$) + modification parameters $\mathcal{U}(-0.20, 0.20)$ $\rightarrow$ $\mathbf{TIF}^*$

3. Considering $\mathbf{TIF}^*$ at Step 2 and item parameters at Step 1:

      - Bruto  $\rightarrow$ Systematically tests

      - ILA  $\rightarrow$ Forwardly searches considering a single $\theta$
      - Frank $\rightarrow$ Forwardly searches considering the whole latent trait
